Rule Synthesis for Rule-based Intrusion Detectors

### 2020-06-02 paper/paper.tex
polishing

%\vspace{0.5ex}
\noindent\textbf{Attacks.}~Table~\ref{table:attacks} shows the selection
of attacks we analyzed.  Non-http attacks are listed at the top of the
table and http attacks are listed at the bottom. Recall from
Section~\ref{sec:reverse-engineering} that (1) http attacks are
prevalent, (2) most http attacks use only content options, and (3)
most non-http attacks only use field options. We selected
\numNonContentAttacks\ non-http attacks and \numContentAttacks\ http
attacks. The higher number of http attacks reflects their
popularity. Column ``Name'' shows the name of the attack, column
``Source'' shows the source where we obtained data to reproduce the
attack (see Section~\ref{subsec:malicious-traffic}), column ``Type''
shows the kind of attack, column ``Protocol'' shows the name of the
protocol explored in the attack, and column ``Ground Truth'' shows the
rule options to capture the attack. The rationale for the selection
were (1) to cover various kinds of attacks and (2) to cover attacks of
various protocols. Considering (1), the selection includes attacks of
the following types: Denial-of-Service (DoS)~\cite{denial-of-service},
Active Reconnaissance (AREC)~\cite{active-reconnaissance}, Privilege
escalation (PESC)~\cite{privilege-escalation}, Remote Code Execution
(RCE)~\cite{remote-code-execution}, Local File Inclusion
(LFI)~\cite{local-file-inclusion} and ICMP Traffic Alerts \cite{icmp-types}.
\noindent\textbf{Malicious traffic.}~Recall that \tname\ takes
pcap files on input. As explained on Section~\ref{sec:minimization},
these files abstract network messages~\cite{pcap} and can be used to
avoid network traffic during rule synthesis. For some of the attacks
listed on Table~\ref{table:attacks}, we obtained pcap files directly
from trusted security websites (\eg{},
NetReseC~\cite{pcap-attacks}). In addition to these sources, we used
the Nikto2~\cite{nikto} open source security tool as follows to obtain
pcap files for some of the http attacks. We configured an http server
(Apache2) and ran Nikto2 to attack that server. Simultaneously, we ran
WireShark~\cite{wireshark-net-monitor} to record the network traffic
and Suricata to analyze the traffic. We configured \suri\ with all
rules from the ``Emerging Threats Open
Ruleset''~\cite{emerging-threats-open} and configured Nikto2 with
default parameters, which makes it replay all recorded attacks from
its benchmark. After execution, we inspected the \suri\ log to
identify which rules have been activated and which packets were
involved on each activation. With the packet numbers, it is possible
to recover the corresponding pcap files, using WireShark, as to ran
\tname{} on them separately. It is worth noting that some rules are
triggered multiple times, indicating that Nikto2 expresses multiple
variants of the same attack. RQ3 evaluates the importance of these
variants to \tname{}'s performance. We used similar methodology to
obtain pcap files using other security testing tools. For example, we
used NMAP~\cite{netmap} to reproduce the Ping Scan attack (see
Section~\ref{sec:active-recon}) and hping3~\cite{hping3} to reproduce
the SYN Flood attack (see Section~\ref{sec:dos}).
\noindent\textbf{Benign traffic.}~Public data sets of benign traffic
exist and can be used to evaluate the rate of false positives of
security tools. We used two popular public sources. The Bigflows.pcap
data set is part of the TcpReplay~\cite{tcpreplay} open source
project. It includes real network traffic of a private network. We
also used data sets from the Stratosphere IPS
Project~\cite{stratosphere-normal} including, for example, the regular
traffic of a university network.
\noindent\textbf{Ground Truth.}~To objectively evaluate \tname{}'s
performance we compared the rules it generates with \emph{trusted
  rules}. We used the ``Emerging Threats Open
Ruleset''~\cite{emerging-threats-open} for that. After finding a
matching pair of attack and corresponding rule using the procedure
described in Section~\ref{subsec:malicious-traffic}, we reproduced the
attack to confirm the rule captures the attack in isolation. Column
``Ground Truth'' from Table~\ref{table:attacks} only shows the
essential options. Documentation options (\eg{}, ``msg'') are
discarded.

@@

Table~\ref{table:weights} shows the results, where ``Name'' is the
name of the attack and ``base'' is the result of
\tname\ with all heuristic functions enabled, as in ``Rank
(step 3)'' in Table~\ref{table:results}. The remaining columns show
results obtained by running the same experiment, as on
Section~\ref{sec:answer-rqone}, but suppressing one heuristic
function. The number corresponds to the order the heuristic function
is introduced on Section~\ref{sec:ranking}. The symbol of an up arrow
($\uparrow$) indicates that the performance of \tname{} on that
configuration is worse when compared to ``base''. A down arrow ($\downarrow$), 
highlighted in the table, indicates the contrary.

@@

%\subsubsection{Lack of comparison with other tools}
\noindent\textbf{Lack of comparison with other tools}
Table~\ref{table:comparison} summarizes the characteristics of
alternative techniques and provides details about the evaluation used
in corresponding papers. Unfortunately, side comparison was not
possible for two central reasons. First, we did not have access to the
tools. We tried, unsuccssefully, making contact with authors to obtain
the tools. Second, the tools output different objects. For that
reason, a comparison based on published results was not realistic. It
is important to recall that we did not report precision and recall as
it is 100\% always. \tname captures, by construction, all positive
traffic and does not capture any negative traffic. We used a stronger
criteria instead---the position of the correct rule in the ranking. It
is possible that other rules, higher in the ranking, are useful, but
only domain experts can identify that.
\noindent\textbf{Running times.}~We ran all our tests using an
Intel(R) Xeon(R) CPU E3-1240 v6 and 8GB RAM. For attacks with only
categorical and numerical options, the average running time was 6
seconds per attack. For attacks with the content option, the average
running time was 1,445 seconds (=24.1 minutes).  \textbf{Threats to
  Validity.}~Considering external validity. Our results cannot be
generalized beyond the subjects we selected. To mitigate that threat
we selected a variety of attacks with different
characteristics. Unfortunately, we could not access tools for related
techniques, but there is strong evidence that \tname\ performs very
well and is the only technique to support content options, which are
very important in this domain. Considering internal validity, in
principle, the implementation of \tname\ could contain errors. We
mitigated that threat by carefully inspecting code and
results. Considering construct validity, the choice of ranking to
evaluate the approach may not be useful in practice. We considered the
use of ranking of the golden rule a reasonable choice as the approach
places the golden rule close to the top in most cases.

@@

\textbf{Alternative methods for rule synthesis.} There are important
differences between our work and prior work. Vollmer
\etal{}~\cite{vollmer-etal-cics2011} proposed the use of search to
derive rules for Snort, which are very similar to \suri's.  Their
solution focused on icmp attacks for which plausible attacks can be
created directly from packets (see
Section~\ref{sec:rules-and-packets}). They modeled every field of the
icmp packet in chromosomes of candidate solutions (\ie{}, a rule) of a
genetic search. The initial population is a set of candidate solutions
that randomly combine options, modelling the packet fields. They
showed that they were able to generate rules without false negatives
and only 3 false positives across 33K benign packets. The key
differences between our work and Vollmer~\etal{}'s are 1) the support
of different protocols and 2) the use of ranking to discriminate
alternative plausible rules. As for 1, our approach supports a variety
of protocols and options (\eg{}, options that refer to message
payloads). The icmp rules they generate only refer to message fields.
\MyComment{\Luc{eh possivel incluir contents em regras icmp tb. o q a
    secao 2.3 diz q \textbf{geralmente} as regras nao-http possuem
    apenas field options}. } As for 2, our solution additionally ranks
plausible options, which contain \emph{no} false positives or
negatives with respect to the benign and mallicious traffic
considered. Guruprasad~\etal{}~\cite{guruprasad} also used an
evolutionary approach to solve the problem. Although the solution of
Guruprasad~\etal{} supports more protocols than Vollmer~\etal{}', our
approach still supports a larger set of protocols and the ability to
ranks rules. Muthuregunathan~\etal{}~\cite{muthuregunathan-etal2009}
propose a clustering method to create rules. They characterize
candidate solutions as a set of field values and use a clustering
algorithm---hybrid k-means partitioned around menoids---to isolate
malicious from benign traffic. Subsequently, they optimize the formed
clusters using a genetic algorithm and hill-climbing search. Ou
approach differs from theirs in the method used, results, and support
for content options. Kao~\etal~\cite{Kao2015AutomaticNR} proposed
ARGS, a system to synthesize http rules that take malicious traffic as
input. ARGS looks for a common pattern across the different inputs to
define rules. It does not take existing rules as input or benign
traffic. Consequently, false positive can be produced and need to be
handled manually by users. Our approach, in contrast, does not
generate false positives w.r.t. the input traffic. Fallahi
\etal{}\cite{fallahi} proposed flow-based synthesis approach to rule
generation. Instead of analyzing packets individually, their algorithm
investigates and creates the rules for flows in the network. The
algorithm was tested using the ISCX 2012 dataset and showed good
results in terms of precision, but it is limited to that type of
attack. As it won't analyse individual details of each packet, it
wouldn't be able to identify and generate rules for attacks that
require detailed options, such as the HTTP PESC, RCE and LFI.
\textbf{Rule repair.}~Hallahan~\etal{}~\cite{8102263} propose to
repair firewall rules based on false positive detected by the
rules. Although firewall and \nids\ rules have differences and the
problem addressed is different---repair not synthesis---the work is
similar in spirit and inspiring. We remain to evaluate how \tname\ can
be used for repair.  Ganesan \etal{}\cite{ganesan} applied similar
ideas in the context of IDS, \ie{}, they proposed to adjust existing
rules with the goal of capturing variations of existing attacks. They
used the community edition of snort rules from 2012 for the
modifications and tested them with pcap file from the MACCDC 2012
dataset. Their technique is complementary to \tname{}.
\textbf{Anomaly-based intrusion detection.}
Cordy~\etal{}\cite{cordy-etal-issta19} used genetic search to not only
to generate attacks capable of penetrating the tested anomaly-based
NIDS, but also to repair the NIDS to identify these
attacks. \tname\ and their approach are complementary as anomaly-based
and rule-based \nids\ are complementary (see
Section~\ref{sec:background}).

>>>>

%\vspace{0.5ex}
\noindent\textbf{Attacks.}~Table~\ref{table:attacks} shows the selection
of attacks we analyzed.  Non-http attacks are listed at the top of the
table and http attacks are listed at the bottom. Recall from
Section~\ref{sec:reverse-engineering} that (1) http attacks are
prevalent, (2) most http attacks use only content options, and (3)
most non-http attacks only use field options. We selected
\numNonContentAttacks\ non-http attacks and \numContentAttacks\ http
attacks. The higher number of http attacks reflects their
popularity. Column ``Name'' shows the name of the attack, column
``Source'' shows the source where we obtained data to reproduce the
attack (see Section~\ref{subsec:malicious-traffic}), column ``Type''
shows the kind of attack, column ``Protocol'' shows the name of the
protocol explored in the attack, and column ``Ground Truth'' shows the
rule options to capture the attack. The rationale for the selection
were (1) to cover various kinds of attacks and (2) to cover attacks of
various protocols. Considering (1), the selection includes attacks of
the following types: Denial-of-Service (DoS)~\cite{denial-of-service},
Active Reconnaissance (AREC)~\cite{active-reconnaissance}, Privilege
escalation (PESC)~\cite{privilege-escalation}, Remote Code Execution
(RCE)~\cite{remote-code-execution}, Local File Inclusion
(LFI)~\cite{local-file-inclusion} and ICMP Traffic Alerts \cite{icmp-types}.

\noindent\textbf{Malicious traffic.}~Recall that \tname\ takes
pcap files on input. As explained on Section~\ref{sec:minimization},
these files abstract network messages~\cite{pcap} and can be used to
avoid network traffic during rule synthesis. For some of the attacks
listed on Table~\ref{table:attacks}, we obtained pcap files directly
from trusted security websites (\eg{},
NetReseC~\cite{pcap-attacks}). In addition to these sources, we used
the Nikto2~\cite{nikto} open source security tool as follows to obtain
pcap files for some of the http attacks. We configured an http server
(Apache2) and ran Nikto2 to attack that server. Simultaneously, we ran
WireShark~\cite{wireshark-net-monitor} to record the network traffic
and Suricata to analyze the traffic. We configured \suri\ with all
rules from the ``Emerging Threats Open
Ruleset''~\cite{emerging-threats-open} and configured Nikto2 with
default parameters, which makes it replay all recorded attacks from
its benchmark. After execution, we inspected the \suri\ log to
identify which rules have been activated and which packets were
involved on each activation. With the packet numbers, it is possible
to recover the corresponding pcap files, using WireShark, as to ran
\tname{} on them separately. It is worth noting that some rules are
triggered multiple times, indicating that Nikto2 expresses multiple
variants of the same attack. RQ3 evaluates the importance of these
variants to \tname{}'s performance. We used similar methodology to
obtain pcap files using other security testing tools. For example, we
used NMAP~\cite{netmap} to reproduce the Ping Scan attack (see
Section~\ref{sec:active-recon}) and hping3~\cite{hping3} to reproduce
the SYN Flood attack (see Section~\ref{sec:dos}).

\noindent\textbf{Benign traffic.}~Public data sets of benign traffic
exist and can be used to evaluate the rate of false positives of
security tools. We used two popular public sources. The Bigflows.pcap
data set is part of the TcpReplay~\cite{tcpreplay} open source
project. It includes real network traffic of a private network. We
also used data sets from the Stratosphere IPS
Project~\cite{stratosphere-normal} including, for example, the regular
traffic of a university network.

\noindent\textbf{Ground Truth.}~To objectively evaluate \tname{}'s
performance we compared the rules it generates with \emph{trusted
  rules}. We used the ``Emerging Threats Open
Ruleset''~\cite{emerging-threats-open} for that. After finding a
matching pair of attack and corresponding rule using the procedure
described in Section~\ref{subsec:malicious-traffic}, we reproduced the
attack to confirm the rule captures the attack in isolation. Column
``Ground Truth'' from Table~\ref{table:attacks} only shows the
essential options. Documentation options (\eg{}, ``msg'') are
discarded.

@@

The results are in Table~\ref{table:weights}, where ``Name'' is the
name of the attack and ``base'' is the result of
\tname\ with all heuristic functions enabled, as in ``Rank
(step 3)'' in Table~\ref{table:results}. The remaining columns are the
results when suppressing one heuristic
function. The number corresponds to the order of hte heuristic function
as introduced on Section~\ref{sec:ranking}. The up arrow
($\uparrow$) indicates that the performance is worse when compared to ``base'',
whereas, a down arrow ($\downarrow$), 
highlighted in the table, indicates the contrary.

@@

%\subsubsection{Lack of comparison with other tools}

\noindent\textbf{Lack of comparison with other tools}
Table~\ref{table:comparison} summarizes the characteristics of
alternative techniques and provides details about the evaluation used
in corresponding papers. Unfortunately, side comparison was not
possible for two central reasons. First, we did not have access to the
tools. We tried, unsuccssefully, making contact with authors to obtain
the tools. Second, the tools output different objects. For that
reason, a comparison based on published results was not realistic. It
is important to recall that we did not report precision and recall as
it is 100\% always. \tname captures, by construction, all positive
traffic and does not capture any negative traffic. We used a stronger
criteria instead---the position of the correct rule in the ranking. It
is possible that other rules, higher in the ranking, are useful, but
only domain experts can identify that.

\noindent\textbf{Running times.}~We ran all our tests using an
Intel(R) Xeon(R) CPU E3-1240 v6 and 8GB RAM. For attacks with only
categorical and numerical options, the average running time was 6
seconds per attack. For attacks with the content option, it was 
1,445 seconds (=24.1 minutes) on average.

\noindent\textbf{Threats to Validity.}~Considering external validity, 
our results cannot be generalized beyond the subjects we selected. To 
mitigate that threat we selected a variety of attacks with different
characteristics. Unfortunately, related techniques are unavailable, 
but there is strong evidence that \tname\ performs very
well and is the only technique to support content options, an important 
aspect this domain. Considering internal validity, in
principle, the implementation of \tname\ could contain errors. We
mitigated that threat by carefully inspecting code and
results. Considering construct validity, the choice of ranking to
evaluate the approach may not be useful in practice. We considered the
use of ranking of the golden rule a reasonable choice as the approach
places the golden rule close to the top in most cases.

@@

\noindent
\textbf{Alternative methods for rule synthesis.} There are important
differences between our work and prior work. Vollmer
\etal{}~\cite{vollmer-etal-cics2011} proposed the use of search to
derive rules for Snort, which are very similar to \suri's.  Their
solution focused on icmp attacks for which plausible attacks can be
created directly from packets (see
Section~\ref{sec:rules-and-packets}). They modeled every field of the
icmp packet in chromosomes of candidate solutions (\ie{}, a rule) of a
genetic search. The initial population is a set of candidate solutions
that randomly combine options, modelling the packet fields. They
showed that they were able to generate rules without false negatives
and only 3 false positives across 33K benign packets. The key
differences between our work and Vollmer~\etal{}'s are 1) the support
of different protocols and 2) the use of ranking to discriminate
alternative plausible rules. As for 1, our approach supports a variety
of protocols and options (\eg{}, options that refer to message
payloads). The icmp rules they generate only refer to message fields.
\MyComment{\Luc{eh possivel incluir contents em regras icmp tb. o q a
    secao 2.3 diz q \textbf{geralmente} as regras nao-http possuem
    apenas field options}. } As for 2, our solution additionally ranks
plausible options, which contain \emph{no} false positives or
negatives with respect to the benign and mallicious traffic
considered. Guruprasad~\etal{}~\cite{guruprasad} also used an
evolutionary approach to solve the problem. Although the solution of
Guruprasad~\etal{} supports more protocols than Vollmer~\etal{}', our
approach still supports a larger set of protocols and the ability to
ranks rules. Muthuregunathan~\etal{}~\cite{muthuregunathan-etal2009}
propose a clustering method to create rules. They characterize
candidate solutions as a set of field values and use a clustering
algorithm---hybrid k-means partitioned around menoids---to isolate
malicious from benign traffic. Subsequently, they optimize the formed
clusters using a genetic algorithm and hill-climbing search. Ou
approach differs from theirs in the method used, results, and support
for content options. Kao~\etal~\cite{Kao2015AutomaticNR} proposed
ARGS, a system to synthesize http rules that take malicious traffic as
input. ARGS looks for a common pattern across the different inputs to
define rules. It does not take existing rules as input or benign
traffic. Consequently, false positive can be produced and need to be
handled manually by users. Our approach, in contrast, does not
generate false positives w.r.t. the input traffic. Fallahi
\etal{}\cite{fallahi} proposed flow-based synthesis approach to rule
generation. Instead of analyzing packets individually, their algorithm
investigates and creates the rules for flows in the network. The
algorithm was tested using the ISCX 2012 dataset and showed good
results in terms of precision, but it is limited to that type of
attack. As it won't analyse individual details of each packet, it
wouldn't be able to identify and generate rules for attacks that
require detailed options, such as the HTTP PESC, RCE and LFI.

\noindent
\textbf{Rule repair.}~Hallahan~\etal{}~\cite{8102263} propose to
repair firewall rules based on false positive detected by the
rules. Although firewall and \nids\ rules have differences and the
problem addressed is different---repair not synthesis---the work is
similar in spirit and inspiring. We remain to evaluate how \tname\ can
be used for repair.  Ganesan \etal{}\cite{ganesan} applied similar
ideas in the context of IDS, \ie{}, they proposed to adjust existing
rules with the goal of capturing variations of existing attacks. They
used the community edition of snort rules from 2012 for the
modifications and tested them with pcap file from the MACCDC 2012
dataset. Their technique is complementary to \tname{}.

\noindent
\textbf{Anomaly-based intrusion detection.}
Cordy~\etal{}\cite{cordy-etal-issta19} used genetic search to not only
to generate attacks capable of penetrating the tested anomaly-based
NIDS, but also to repair the NIDS to identify these
attacks. \tname\ and their approach are complementary as anomaly-based
and rule-based \nids\ are complementary (see
Section~\ref{sec:background}).
